---
title: Monte-Carlo examples in R
subtitle: Serial vs parallel code
author: Meelis Utt
date: 
output: pdf_document
# output: html_document
header-includes:
 - \usepackage[T1]{fontenc}
 - \usepackage[utf8]{inputenc}
 - \usepackage[estonian]{babel}
 - \usepackage{amsmath}
 - \usepackage{amssymb}
 - \usepackage{amsthm}
 - \usepackage{soul}
 - \usepackage{cancel}
 - \usepackage{enumitem}
 - \usepackage{graphicx}
 - \usepackage{pdfpages}
 - \usepackage{titling}
 - \usepackage[nottoc,numbib]{tocbibind}
 - \usepackage{hyperref}
---

\def\UrlBreaks{\do\/\do-}
\hypersetup{ colorlinks=true, linkcolor=blue, filecolor=magenta,      urlcolor=cyan, citecolor=red}
\newcommand{\refr}[1]{ (\ref{#1}) }
\newcommand{\comment}[2]{ \overset{\makebox[20pt]{\mbox{\normalfont\tiny #1}}}{#2} }

# Setup

```{r setup,include=F,echo=T}
# Check if necessary packages exists.
# If not then download them
check_packages <- function(pkg) {
 new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
 if (length(new.pkg))
  install.packages(new.pkg, dependencies  =  TRUE,repos = "https://ftp.acc.umu.se/mirror/CRAN/") 
 sapply(pkg, library, character.only  =  TRUE)
}

# The following packages are dependencies for this app.
packages <- c("magrittr", "data.table", "ggplot2", "parallel","foreach","doParallel","dplyr")
check_packages(packages)

# # load necessary package
# library(magrittr)
# library(data.table)
# library(ggplot2)
# library(dplyr)
# # library(knitr)
# library(parallel)
# library(foreach)
# library(doParallel)

knitr::opts_chunk$set(fig.width = 6, fig.heighta = 4)
options(scipen = 1000)
```

Let's source the setup (function, analytical solution, number of iterations).

```{r}
source("Setup.R",echo = T)
```

# Serial implementations

Let's start with a simple implementation of Monte-Carlo method.

```{r}
MCser1 <- function(n){
  start <- Sys.time()
  i <- runif(n,0,1)
  EX <- mean(f(i))
  end <- Sys.time()
  time <- difftime(end,start)
  error <- (EX - analytical) %>% abs
  return(c(n,EX,analytical,error,time,"MCser1"))
}
data.table(t(MCser1(n))) %>% setNames(header)
```

Let's try a bit more vectorized solution, using the \*apply function.

```{r}
MCser2 <- function(n,ncols=1000){
  start <- Sys.time()
  dt <- matrix(runif(n,0,1),ncol = ncols)
  EX <- sapply(1:ncols,function(i,dt){
      EX <- dt[,i] %>% f %>% mean
    },dt) %>% mean
  end <- Sys.time()
  time <- difftime(end,start)
  error <- (EX - analytical) %>% abs
  return(c(n,EX,analytical,error,time,"MCser2"))
}
data.table(t(MCser2(n))) %>% setNames(header)
```

Let's try an approach using data.table.

```{r}
MCser3 <- function(n){
  start <- Sys.time()
  dt <- data.table(unif = runif(n,0,1))
  EX <- dt[,.(EX = mean(f(unif)))] %>% unlist %>% unname
  end <- Sys.time()
  time <- difftime(end,start)
  error <- (EX - analytical) %>% abs
  return(c(n,EX,analytical,error,time,"MCser3"))
}
data.table(t(MCser3(n))) %>% setNames(header)
```

Let's try divide-and-conquer approach with data.table.

```{r}
MCser4 <- function(n,ncols=1000){
  start <- Sys.time()
  dt <- matrix(runif(n,0,1),ncol=ncols) %>% data.table
  EX <- dt[,lapply(.SD,f)][,lapply(.SD,mean)][,.(EX = sum(.SD)/ncols)] %>% unlist %>% unname
  end <- Sys.time()
  time <- difftime(end,start)
  error <- (EX - analytical) %>% abs
  return(c(n,EX,analytical,error,time,"MCser4"))
}
data.table(t(MCser4(n))) %>% setNames(header)
```


# Parallel implementations

Let's try different parallel implementations.
First let's start with package \textit{parallel}.

```{r}
MCpar1 <- function(n){
  start <- Sys.time()
  # Calculate the number of cores
  no_cores <- detectCores()
  # Initiate cluster
  cl <- makeCluster(no_cores)
  intermean <- parSapply(cl, rep(n/no_cores,no_cores),function(ni,f){
      EX <- mean(f(runif(ni,0,1)))
    },f
  )
  on.exit(stopCluster(cl))
  EX <- mean(intermean)
  end <- Sys.time()
  time <- difftime(end,start)
  error <- (EX - analytical) %>% abs
  return(c(n,EX,analytical,error,time,"MCpar1"))
}
data.table(t(MCpar1(n))) %>% setNames(header)
```

Now let's try approach analogical to MCser2.

```{r}
MCpar2 <- function(n){
  start <- Sys.time()
  # Calculate the number of cores
  no_cores <- detectCores()
  cl <- makeCluster(no_cores)
  dt <- matrix(runif(n,0,1),ncol = no_cores)
  intermean <- parSapply(cl, 1:no_cores,function(i,f,dt){
      EX <- mean(f(dt[,i]))
    },f,dt
  )
  on.exit(stopCluster(cl))
  EX <- mean(intermean)
  end <- Sys.time()
  error <- (EX - analytical) %>% abs
  time <- difftime(end,start)
  return(c(n,EX,analytical,error,time,"MCpar2"))
}
data.table(t(MCpar2(n))) %>% setNames(header)
```

This approach was not very good. But let's have one more try at analogical solution to MCser2.

```{r}
MCpar2_2 <- function(n,ncols=1000){
  start <- Sys.time()
  # Calculate the number of cores
  no_cores <- detectCores()
  cl <- makeCluster(no_cores)
  dt <- matrix(runif(n,0,1),ncol = ncols)
  intermean <- parSapply(cl, 1:ncols,function(i,f,dt){
      EX <- mean(f(dt[,i]))
    },f,dt
  )
  on.exit(stopCluster(cl))
  EX <- mean(intermean)
  end <- Sys.time()
  error <- (EX - analytical) %>% abs
  time <- difftime(end,start)
  return(c(n,EX,analytical,error,time,"MCpar2_2"))
}
data.table(t(MCpar2_2(n))) %>% setNames(header)
```

This solution was bit better, but still worse than the previous examples.\
Let's try the package \textit{foreach} now.

```{r}
MCpar3 <- function(n){
  start <- Sys.time()
  # Calculate the number of cores
  no_cores <- detectCores()
  # Initiate cluster
  cl<-makeCluster(no_cores)
  # registerDoParallel(cl)
  EX <- foreach(ni = rep(n/no_cores,no_cores),.combine=mean,.export="f") %dopar% 
    mean(f(runif(ni,0,1)))
  on.exit(stopCluster(cl))
  # stopImplicitCluster()
  end <- Sys.time()
  time <- difftime(end,start)
  error <- (EX - analytical) %>% abs
  return(c(n,EX,analytical,error,time,"MCpar3"))
}
data.table(t(MCpar3(n))) %>% setNames(header)
```


# Benchmarking

Now, let's visualize the walltimes of implemented solutions.

```{r visualization}
iterations <- c(2.5,5,7.5,10)*10**(5:7)
funs <- c(
  MCser1,MCser2,MCser3,MCser4,
  MCpar1,MCpar3 #MCpar2,MCpar2_2,
)
data <- sapply(funs,function(f,iterations){
  sapply(iterations,function(n){
    f(n)
  })
},iterations) %>% 
  matrix(ncol=6,byrow=T) %>%
  data.table %>%
  setNames(header) %>%
  mutate_at(header[-grep(x=header,pattern="type")],as.numeric)

data

ggplot(data=data,aes(x=n,y=walltime,group=type,color=type)) +
  geom_point() +
  geom_line() +
  labs(
       title="Serial vs parallel implementations",
       x="Nr of iterations",
       y="Walltime (sec)"
       ) +
  theme(
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)
       ) +
  guides(color=guide_legend(title="Function"))

```

